Exploring probability distributions for bivariate temporal granularities

### 2019-12-04 CyclicGran.Rmd
spell check done

where,  
$z$ denotes the index set,
$B$ denotes a full-integer labelled bottom granularity which groups periodicially into linear granularity $G'$ with irregular mapping, {L} is a label mapping that defines an unique label for each index $l \in {0,1,\dots, (k-1)}$, $T_w$ are the sets of indices of B describing $G'(w)$ such that 
\begin{equation}
G'(w) = \bigcup_{z \in T_w}B(z),\\
\end{equation}
$T_w = 0 \quad for \quad w\epsilon Z_{<0}$ and $\vert T_w \vert$ is the cardinality of set $T_w$.
\end{definition}

>>>>

where,  
$z$ denotes the index set,
$B$ denotes a full-integer labelled bottom granularity which groups periodically into linear granularity $G'$ with irregular mapping, {L} is a label mapping that defines an unique label for each index $l \in {0,1,\dots, (k-1)}$, $T_w$ are the sets of indices of B describing $G'(w)$ such that 
\begin{equation}
G'(w) = \bigcup_{z \in T_w}B(z),\\
\end{equation}
$T_w = 0 \quad for \quad w\epsilon Z_{<0}$ and $\vert T_w \vert$ is the cardinality of set $T_w$.
\end{definition}

### 2019-12-04 main.Rmd
spell check done

**Dicussion:** The first condition in \ref{def:linear} implies that the granules in a linear granularity are non-overlapping and their index order is same as time order.
\autoref{fig:linear-time} shows the implication of this condition. It shows the linear granularities hours, days, weeks, months and years, each of which are unidirectional in nature and arranged from past to future. If we consider the chronon [@aigner2011visualization] as hourly, the time domain with T hours will have $\lfloor T/24\rfloor$ days, $\lfloor T/(24*7)\rfloor$ weeks and so on. Each granule of the linear granularities shown are represented through a box.

@@

**Dicussion:** $G \trianglelefteq H$ if each granule H(j) is the union of some granules of G.  $G \preceq H$ if every granule in G is a subset of some granule in H. In \autoref{fig:groups-finer}, both $Days \trianglelefteq Weeks$ and $Days \preceq Weeks$ hold true. However, Weekends groups into Holidays but Weekends are not finer than Holidays. On the other hand, Weekends are finer than weeks but do not group into weeks. Consider another example where $Days \trianglelefteq Months$. This relationship is not fully described until we associate it with periodicity. Each month is a grouping of the same number of days over years, hence the period of the grouping (day, month) is one year, if leap years are ignored. The period becomes 400 years with the inclusion of leap years and all their exceptions, .

@@

**Dicussion:**

@@

**Dicussion:** Given the set of all granularities defined over the time domain (Z; <), and the granularity relationship $\trianglelefteq$ (groups into), the granularity mapping each index into the corresponding instant (same integer number as the index) is a bottom granularity with respect to $\trianglelefteq$.

@@

For more variations of grouping operations and granule oriented operations, the readers are remcommended to see [@Ning2002-tf].s

@@

We propose a formalism of cyclic time granularities through the tsibble [@wang2019tsibble] framework of organizing temporal data. A time domain, as defined by [@Bettini1998-ed], is essentially a mapping of the index set to the time index of a tsibble. A linear granularity is a mapping of the index set to subsets of the time domain. For example, if the time index is days, then a linear granularity might be weeks, months or years. Lining up with the the assumption in [@Bettini2000-vy] that all linear granularities can be generated from the bottom granularity by calendar algebraic operations, we assume that a bottom granularity exists and is represented by the index of the tsibble. In a cyclic organization of time, the domain is composed of a set of recurring time values, where any time value can be preceeded and succeeded by another time value (for e.g Monday comes before Wednesday but Monday also succeeds Wednesday). Hence intuitively, cyclic granularities are additional abstractions of time which are not linear and hence the index order of linear granularities needs to be manipulated so that it leads to repetitive categorization of time.

Cyclic time granularities which accomodate periodicities can be constructed by relating two linear granularities. The mappings between linear granularities can be regular or irregular. For example, a regular mapping exists between minutes and hours, where 60 minutes always add up to 1 hour. On contrary, an irregular mapping exists between days and months, since one might can have days ranging from 28 to 31. To elaborate more, mappings will be regular if, for example, linear granularities are formed by grouping operation. Grouping operations combine fixed granules of a linear granularity to form a new granule of the resulting linear granularity. However, if granularities are formed by the altering-tick operation, for example, granules of the resulting granularity is composed of different number of granules of the root granularity. Cyclic time granularities are referred to as **circular** if mappings between linear time granularities are regular and **quasi-circular** if the same is irregular. Examples of circular granularities include hour of the day, and day of the week, whereas examples for quasi-circular granularities can be day of the month or week of the month. The formalisms would differ for circular and quasi-circular granularities although both accomodate periodicities in data.

@@

\begin{equation} \label{eq:eq2}
C_{B, G}(z) = {L}(z\mod k(B,G)) \quad \forall z \in \mathbb{Z}^+
\end{equation}
where,\\
z denotes the index set, \\
B denotes a full-integer labelled bottom granularity which groups periodicially into linear granularity G with regular mapping, \\
{L} is a label mapping that defines an unique label for each index $l \in {0,1,\dots, (k-1)}$,\\
and k(B,G) is the number of granules of B in each granule of G.
\end{timedom}

**Discussion**:  In general, the set of labels will be a set of strings that is more descriptive than the index and used to identify a categorisation of the circular granularity. However, the labels can coincide with indexes in which case integers are directly used to refer to categorisations of the circular granularity. Note that each
circular granularity can use a different set of labels. Example showing circular granularities relating two linear granularities each with bottom granularities are visually depicted in a series of slots in the \autoref{fig:circular-gran}. Each granule is represented by a box. The diagram also illustrates that the granules that overlap share elements from the underlying time domain. The first slot in the diagram shows the index set and the bottom granularity. The index of the tsibble considered is days. G and H are two linear granularities denoting days and weeks respectively. $k(day, week) =  7$ and hence, $C_{G, B}(z)$ represented by C(BG, G) is given by $B(z\mod 7)$. The circular granularity C(B, G) consists of repetitive pattern {L(0), L(1), L(2), L(3), L(4), L(5), L(6)} which repeats itself after each period length. $L$ can be defined such that the labels corresponding to ${0,1,\dots,6}$ correspond to different days of the week ${Sun, Mon ,\dots, Sat}$ . Similarly, C(B, H) is given by $BG(z\mod 14)$ since the period length of days with respect to fortnight is 14. C(B, H) also repeats the pattern ${L'(0), \dots, L'(13)}$ in every period. Hence, circular granularities are represented as repetitive categorization of time in the diagram.

@@

where,  
$z$ denotes the index set,\\
$B$ denotes a full-integer labelled bottom granularity which groups periodicially into linear granularity $G'$ with irregular mapping,\\
{L} is a label mapping that defines an unique label for each index $l \in {0,1,\dots, (k-1)}$,\\
$T_w$ are the sets of indices of B describing $G'(w)$ such that 
\begin{equation}
G'(w) = \bigcup_{z \in T_w}B(z),\\
\end{equation}
$T_w = 0$ for $w \epsilon Z_{<0}$,\\
$\vert T_w \vert$ is the cardinality of set $T_w$.
\end{timedom}

@@

<!-- The cyclic time granularities which are essentially categorisations of time are discrete variable, mapped as a factor since they have an inherent order. Since two cyclic granularities are mapped to the the fixed and conditioned variable, the way in which they are related becomes important in the data structure. To define a general framework, we define harmony and clashes. -->

@@

The measured variable can be dicrete or continuous. The probability distribution of these univariate measured variable could be studied by estimating the proability density function (pdf for continuous) or probability mass function (pmf for dicrete). Both pdf and pmf are concerned with point probabilities of the random variable. Suppose we have observations which are realizations of univariate random variables, $X_1, X_2, \dots, X_n \sim F$, where $F$ denotes an unknown cumulative distribution function and $X_1, X_2, \dots, X_n$ are i.i.d. The densities can either be estimated or summarized using quantiles.Instead of assuming a parametric model for the distribution (e.g. a Normal distribution with unknown expectation and variance), a non-parametric estimation assumes that the density exists and is suitably smooth (e.g. differentiable). It is then possible to estimate the unknown density function f(·).

The histogram is the oldest and most popular density estimator method. The form of the histogram depends very much on the tuning parameters class width and origin. The naive estimator was introduced by Fix and Hodges (1951) and first discussed by Rosenblatt(1956). Whittle (1958) formulated the general weight function class of estimators, while the orthogonal series estimator was introduced by Cencov (1962). The nearest neighbour estimate considered by Loftsgaarden and Quesenbury (1965), while the variable Kernel method is due to Brieman, Meisel and Purcell (1977). There are quite a few other approaches to density estimation than the kernel estimators, but the fixed bandwidth kernel estimators are used predominantly because of their simplicity.

@@

There have been several other nonparametric quantile estimators proposed e.g., Harrell and Davis 1982; Sheather and Marron 1990). Also sample quantiles can be of the form that are not defined for all p including hinges, letter values (Hoaglin 1983) and related methods (Freund and Perles 1987).

<!-- would invlove a faceting approach with one temporal granularity plotted along the x-axis, the other one across facet and the dependent time series variable on the y-axis. -->

>>>>

**Discussion:** The first condition in \ref{def:linear} implies that the granules in a linear granularity are non-overlapping and their index order is same as time order.
\autoref{fig:linear-time} shows the implication of this condition. It shows the linear granularities hours, days, weeks, months and years, each of which are unidirectional in nature and arranged from past to future. If we consider the chronon [@aigner2011visualization] as hourly, the time domain with T hours will have $\lfloor T/24\rfloor$ days, $\lfloor T/(24*7)\rfloor$ weeks and so on. Each granule of the linear granularities shown are represented through a box.

@@

**Discussion:** $G \trianglelefteq H$ if each granule H(j) is the union of some granules of G.  $G \preceq H$ if every granule in G is a subset of some granule in H. In \autoref{fig:groups-finer}, both $Days \trianglelefteq Weeks$ and $Days \preceq Weeks$ hold true. However, Weekends groups into Holidays but Weekends are not finer than Holidays. On the other hand, Weekends are finer than weeks but do not group into weeks. Consider another example where $Days \trianglelefteq Months$. This relationship is not fully described until we associate it with periodicity. Each month is a grouping of the same number of days over years, hence the period of the grouping (day, month) is one year, if leap years are ignored. The period becomes 400 years with the inclusion of leap years and all their exceptions, .

@@

**Discussion:**

@@

**Discussion:** Given the set of all granularities defined over the time domain (Z; <), and the granularity relationship $\trianglelefteq$ (groups into), the granularity mapping each index into the corresponding instant (same integer number as the index) is a bottom granularity with respect to $\trianglelefteq$.

@@

For more variations of grouping operations and granule oriented operations, the readers are recommended to see [@Ning2002-tf].s

@@

We propose a formalism of cyclic time granularities through the tsibble [@wang2019tsibble] framework of organizing temporal data. A time domain, as defined by [@Bettini1998-ed], is essentially a mapping of the index set to the time index of a tsibble. A linear granularity is a mapping of the index set to subsets of the time domain. For example, if the time index is days, then a linear granularity might be weeks, months or years. Lining up with the the assumption in [@Bettini2000-vy] that all linear granularities can be generated from the bottom granularity by calendar algebraic operations, we assume that a bottom granularity exists and is represented by the index of the tsibble. In a cyclic organization of time, the domain is composed of a set of recurring time values, where any time value can be preceded and succeeded by another time value (for e.g Monday comes before Wednesday but Monday also succeeds Wednesday). Hence intuitively, cyclic granularities are additional abstractions of time which are not linear and hence the index order of linear granularities needs to be manipulated so that it leads to repetitive categorization of time.

Cyclic time granularities which accommodate periodicities can be constructed by relating two linear granularities. The mappings between linear granularities can be regular or irregular. For example, a regular mapping exists between minutes and hours, where 60 minutes always add up to 1 hour. On contrary, an irregular mapping exists between days and months, since one might can have days ranging from 28 to 31. To elaborate more, mappings will be regular if, for example, linear granularities are formed by grouping operation. Grouping operations combine fixed granules of a linear granularity to form a new granule of the resulting linear granularity. However, if granularities are formed by the altering-tick operation, for example, granules of the resulting granularity is composed of different number of granules of the root granularity. Cyclic time granularities are referred to as **circular** if mappings between linear time granularities are regular and **quasi-circular** if the same is irregular. Examples of circular granularities include hour of the day, and day of the week, whereas examples for quasi-circular granularities can be day of the month or week of the month. The formalism would differ for circular and quasi-circular granularities although both accommodate periodicities in data.

@@

\begin{equation} \label{eq:eq2}
C_{B, G}(z) = {L}(z\mod k(B,G)) \quad \forall z \in \mathbb{Z}^+
\end{equation}
where,\\
z denotes the index set, \\
B denotes a full-integer labelled bottom granularity which groups periodically into linear granularity G with regular mapping, \\
{L} is a label mapping that defines an unique label for each index $l \in {0,1,\dots, (k-1)}$,\\
and k(B,G) is the number of granules of B in each granule of G.
\end{timedom}

**Discussion**:  In general, the set of labels will be a set of strings that is more descriptive than the index and used to identify a categorization of the circular granularity. However, the labels can coincide with indexes in which case integers are directly used to refer to categorizations of the circular granularity. Note that each
circular granularity can use a different set of labels. Example showing circular granularities relating two linear granularities each with bottom granularities are visually depicted in a series of slots in the \autoref{fig:circular-gran}. Each granule is represented by a box. The diagram also illustrates that the granules that overlap share elements from the underlying time domain. The first slot in the diagram shows the index set and the bottom granularity. The index of the tsibble considered is days. G and H are two linear granularities denoting days and weeks respectively. $k(day, week) =  7$ and hence, $C_{G, B}(z)$ represented by C(BG, G) is given by $B(z\mod 7)$. The circular granularity C(B, G) consists of repetitive pattern {L(0), L(1), L(2), L(3), L(4), L(5), L(6)} which repeats itself after each period length. $L$ can be defined such that the labels corresponding to ${0,1,\dots,6}$ correspond to different days of the week ${Sun, Mon ,\dots, Sat}$ . Similarly, C(B, H) is given by $BG(z\mod 14)$ since the period length of days with respect to fortnight is 14. C(B, H) also repeats the pattern ${L'(0), \dots, L'(13)}$ in every period. Hence, circular granularities are represented as repetitive categorization of time in the diagram.

@@

where,  
$z$ denotes the index set,\\
$B$ denotes a full-integer labelled bottom granularity which groups periodically into linear granularity $G'$ with irregular mapping,\\
{L} is a label mapping that defines an unique label for each index $l \in {0,1,\dots, (k-1)}$,\\
$T_w$ are the sets of indices of B describing $G'(w)$ such that 
\begin{equation}
G'(w) = \bigcup_{z \in T_w}B(z),\\
\end{equation}
$T_w = 0$ for $w \epsilon Z_{<0}$,\\
$\vert T_w \vert$ is the cardinality of set $T_w$.
\end{timedom}

@@

<!-- The cyclic time granularities which are essentially categorizations of time are discrete variable, mapped as a factor since they have an inherent order. Since two cyclic granularities are mapped to the the fixed and conditioned variable, the way in which they are related becomes important in the data structure. To define a general framework, we define harmony and clashes. -->

@@

The measured variable can be dicrete or continuous. The probability distribution of these univariate measured variable could be studied by estimating the probability density function (pdf for continuous) or probability mass function (pmf for dicrete). Both pdf and pmf are concerned with point probabilities of the random variable. Suppose we have observations which are realizations of univariate random variables, $X_1, X_2, \dots, X_n \sim F$, where $F$ denotes an unknown cumulative distribution function and $X_1, X_2, \dots, X_n$ are i.i.d. The densities can either be estimated or summarized using quantiles.Instead of assuming a parametric model for the distribution (e.g. a Normal distribution with unknown expectation and variance), a non-parametric estimation assumes that the density exists and is suitably smooth (e.g. differentiable). It is then possible to estimate the unknown density function f(·).

The histogram is the oldest and most popular density estimator method. The form of the histogram depends very much on the tuning parameters class width and origin. The naive estimator was introduced by Fix and Hodges (1951) and first discussed by Rosenblatt(1956). Whittle (1958) formulated the general weight function class of estimators, while the orthogonal series estimator was introduced by Cencov (1962). The nearest neighbor estimate considered by Loftsgaarden and Quesenbury (1965), while the variable Kernel method is due to Brieman, Meisel and Purcell (1977). There are quite a few other approaches to density estimation than the kernel estimators, but the fixed bandwidth kernel estimators are used predominantly because of their simplicity.

@@

There have been several other non parametric quantile estimators proposed e.g., Harrell and Davis 1982; Sheather and Marron 1990). Also sample quantiles can be of the form that are not defined for all p including hinges, letter values (Hoaglin 1983) and related methods (Freund and Perles 1987).

<!-- would involve a faceting approach with one temporal granularity plotted along the x-axis, the other one across facet and the dependent time series variable on the y-axis. -->
